### 2019-09-26 ch7_selfassessment/content.tex
update

The number of datasets used and methods compared against is also below expectations (Figure \ref{fig:benchmarks_over_time}C,D). Only three TI articles feature a comparison of at least 5 methods using 5 datasets or more \cite{sharma_forksfindingorderings_2017,guo_hoplandsinglecellpseudotime_2017,parra_reconstructingcomplexlineage_2018}.

While self-assessments are universally biased in favour of the authors\cite{norel_selfassessmenttrapcan_2011} (intentially or not), it is dangerous and unusual to publish a computational tool without quantitatively demonstrating its performance compared to state-of-the-art methods. Indeed, a recent comparison of 45 TI methods demonstrated that most methods perform worse than a few baseline methods constructed by combining simple off-the-shelf algorithms such as PCA, $k$-means and MST\cite{saelens_comparisonsinglecelltrajectory_2019}.

@@

@@

Several authors use mainly real datasets to evaluate their method, though only few use more than four datasets (n=7). 
By now, already hundreds of suitable real datasets are available from GEO and ArrayExpress (Figure~\ref{fig:datasets}). Downloading and pre-processing them requires a significant time investment, but by processing the datasets once and storing them in a single repository they can be reused for multiple purposes.

Readers are welcome to reuse (and extend) the 110 real and 229 synthetic datasets used in our comparison of TI methods. The datasets are hosted on Zenodo\cite{cannoodt_singlecellomicsdatasets_2018} and the scripts to process them on GitHub\footnote{\href{https://github.com/dynverse/dynbenchmark/tree/master/scripts/01-datasets}{github.com/dynverse/dynbenchmark/tree/master/scripts/01-datasets}}. Note that the ground truths of the datasets are represented using the common data structures format in the previous section.

\section{Multiple metrics}
The largest problem, by far, seems to be a lack of standardised metrics to evaluate TI methods. None of the benchmarks use a metric directly aimed at comparing the topology or pairwise orderings of two trajectories.

Instead, most benchmarks (n=26) employ an ordering metric (e.g. pearson correlation) the pseudotime\footnote{The pseudotime value of is calculated by computing its distance from a pre-defined or inferred start cell.} of the predicted trajectory to ground truth information such as the cell type or time of sampling. Several benchmarks (n=5) use a clustering metric (e.g. ARI) to compare a cell's transition assignment or milestone assignment to the ground truth cell type. In four cases, multiple executions were compared to evaluate the robustness of the predictions.
In one particular instance, the pseudotime correlation metric was adapted to be suitable for comparing rooted trees\cite{street_slingshotcelllineage_2018}. In another, an internal measure is used to quantify the smoothness of gene expression along the pseudotime \cite{darocha_reconstructioncomplexsinglecell_2018}.

While each of these metrics provide some insight the correctness of cell orderings or cell clusterings, they do not evaluate the correctness of the prediction's topology. Using only a pseudotime-based metric to evaluate non-linear trajectories is particularly risky, as it

\begin{figure}[htb!]
	\centering
	\includegraphics[width=.5\linewidth]{fig/metrics.pdf} 
	\caption{}
	\label{fig:metrics}
\end{figure}

\section{Further guidelines}

>>>>

The number of datasets used and methods compared against is also below expectations (Figure \ref{fig:benchmarks_over_time}C,D). Only three TI articles feature a comparison of at least 5 methods using 5 datasets or more \cite{sharma_forksfindingorderings_2017,guo_hoplandsinglecellpseudotime_2017,parra_reconstructingcomplexlineage_2018}. In comparison, our recent benchmark of TI methods evaluated the performance 45 TI methods on 110 real and 229 synthetic datasets\cite{saelens_comparisonsinglecelltrajectory_2019}.

While self-assessments are universally biased in favour of the authors\cite{norel_selfassessmenttrapcan_2011} (intentially or not), it is dangerous and unusual to publish a computational tool without quantitatively demonstrating its performance compared to state-of-the-art methods. Indeed, our comparison demonstrated that most methods perform worse than a few baseline methods constructed by combining simple off-the-shelf algorithms such as PCA, $k$-means and MST.

@@

\mycomment{Todo: mention containerised methods in dynmethods or simply using them as standalone containers.}

@@

Several authors use mainly real datasets to evaluate their method, though only few use more than four datasets (n=7). 
By now, already hundreds of suitable real datasets are available from GEO and ArrayExpress (Figure~\ref{fig:datasets}). Downloading and pre-processing them requires a significant time investment, but by processing the datasets once and storing them in a single repository they can be reused for multiple purposes. Mixture control experiments\cite{tian_benchmarkingsinglecell_2019} are particularly useful in this context.

Readers are welcome to reuse (and extend) the 110 real and 229 synthetic datasets used in our comparison of TI methods\cite{saelens_comparisonsinglecelltrajectory_2019}. The datasets are hosted on Zenodo\cite{cannoodt_singlecellomicsdatasets_2018} and the scripts to process them on GitHub\footnote{\href{https://github.com/dynverse/dynbenchmark/tree/master/scripts/01-datasets}{github.com/dynverse/dynbenchmark/tree/master/scripts/01-datasets}}. Note that the ground truths of the datasets are represented using the common data structures format in the previous section.

\section{Metrics}
To evaluate a TI method, a quantitative metric is needed to compare the predicted trajectory to the ground truth trajectory. No off-the-shelf metrics exist for comparing complex multilayered data structures such as trajectories to each other.
To get around this problem, most benchmarks repurpose metrics from other domains (Figure~\ref{fig:metrics}A).

For example, most benchmarks (n=26) compare the linear pseudotime ordering of a trajectory with ground-truth information such as time time of sampling, quite often by calculating the pearson correlation. This is a good approach for comparing linear trajectories, but is not suitable as a metric for comparing non-linear trajectories (e.g. by calculating the distance from one end point of the trajectory), as this metric does not capture any differences in topology between the two trajectories (Figure~\ref{fig:metrics}B).
Several other benchmarks (n=5) use a metric typically used to compare a clustering method, by comparing a cell's assignment to the transitions of the trajectory to ground-truth information such as the cell's cell type. While this will provide some information on whether cells are grouped correctly in comparison to the ground-truth, it also does not capture topological differences between trajectories (Figure~\ref{fig:metrics}C).

\begin{figure}[htb!]
	\centering
	\includegraphics[width=.6\linewidth]{fig/perturbations.pdf} 
	\caption{
		\textbf{A:} Without any off-the-shelf metrics to use for evaluating TI methods, authors use metrics from other domains.
		\textbf{B:} Comparing these two trajectories using a pseudotime metric (e.g. pearson correlation) would return a perfect score, even though these two trajectories are clearly different.
		\textbf{C:} Comparing these two trajectories using a clustering metric (e.g. ARI) would also return a perfect score, even though these two trajectories are clearly different.
	}
	\label{fig:metrics}
\end{figure}

Extra credit should be given to four cases in which the robustness of methods was evaluated by comparing multiple executions of the same method\cite{haghverdi_diffusionpseudotimerobustly_2016,cannoodt_scorpiusimprovestrajectory_2016,_doesyourcode_2018,sharma_forksfindingorderings_2017,jin_scepathenergylandscapebased_2018}. Computing the robustness does not replace the necessity of a relevant metric that captures whether a predicted trajectory resembles the ground truth -- that is, a TI method can robustly make incorrect predictions and obtain high robustness scores.
In another benchmark, an internal metric is used to quantify the smoothness of gene expression along the pseudotime \cite{darocha_reconstructioncomplexsinglecell_2018} -- the idea being that good TI methods would order cells such that gene expression is smooth along the pseudotime.

In our comparison of TI methods\cite{saelens_comparisonsinglecelltrajectory_2019}, we use a metric called the \emph{geodesic correlation}. Here, two trajectories are compared by calculating the geodesic distances between pairs of cells and comparing those distance values using a Pearson correlation. Note not all pairwise distances are evaluated in this way, or the metric would not scale well to larger datasets. We also use the Hamming-Ipsen-Mikhailov (HIM) distance\cite{jurman_himglocalmetric_2015} to compare the topology of two trajectories.

In Supplementary Note 1 of our comparison of TI methods, we describe and illustrate 10 different metrics, including the geodesic correlation metric, the HIM distance, and several clustering and internal metrics. We constructed 26 test cases (similar to Figure~\ref{fig:metrics}B,C) to assess whether a metric is able of capturing the desired information. We found that the geodesic distance passes nearly all of the test cases, using the geometric mean of different metrics performs much better.

All of these metrics are described in detail in Supplementary Note 1 and implementations are available in \texttt{dyneval}\cite{dyno}.

\section{Guidelines for performing self-assessments}
Articles introducing novel trajectory inference methods have unfortunately been plagued by low self-assessment rates. Most articles passing peer-review do not provide quantitative evidence that their method performs well. In the previous sections, we show that this is most likely caused by differing problem statements making methods hard to compare and a lack of benchmarking datasets and relevant metrics. We provide the reader with viable solutions to each of the problems, in order to be able to benchmark their method with ease.

We wish to leave the reader with a set of guidelines on how to perform a self-assessment on the performance of their own computational method.
%refer to: \cite{norel_selfassessmenttrapcan_2011} 
%In order to alleviate the overestimation of accuracy from the many bias sources described above, we proposed a few guidelines:
%* use third‐party validation to test a model with previously unseen data
%* use more than one metric to evaluate the methods
%* report well‐performing methods even if they are not the best performers on a particular data set
%* increase the awareness of editors and reviewers that superior performance in self‐assessment is a biased demonstration of the method's value; instead, impartial assessment should be the preferred evaluation
%* Establish a scientific culture that values timely, well‐conducted follow‐up studies that confirm or refute previous results

\mycomment{Get a few guidelines from \cite{norel_selfassessmenttrapcan_2011} and \cite{weber_essentialguidelinescomputational_2019}.}

\begin{table}[h]
	\caption{\textbf{Trajectory inference method metadata.} } \label{tab:methods}
	\centering
	\tiny
	\rowcolors{2}{gray!10}{white}
\begin{tabularx}{\linewidth}{|lp{1cm}p{1cm}lXXX|}
	\hline
	id & preprint date & publication date & doi & methods & metrics & datasets \\ \hline \hline
monocle1 &  & 2014-03-23 & \doi{10.1038/nbt.2859} &  &  &  \\
wanderlust &  & 2014-04-24 & \doi{10.1016/j.cell.2014.04.005} &  &  &  \\
scuba &  & 2014-12-30 & \doi{10.1073/pnas.1408993111} &  &  &  \\
sincell & 2015-01-27 & 2015-06-22 & \doi{10.1093/bioinformatics/btv368} &  &  &  \\
nbor &  & 2015-06-08 & \doi{10.1038/ni.3200} &  &  &  \\
oscope &  & 2015-08-24 & \doi{10.1038/nmeth.3549} &  &  &  \\
cycler &  & 2015-08-24 & \doi{10.1038/nmeth.3545} &  &  &  \\
waterfall &  & 2015-09-03 & \doi{10.1016/j.stem.2015.07.013} &  &  &  \\
gpseudotime & 2015-09-15 &  & \doi{10.1101/026872} &  &  &  \\
embeddr & 2015-09-18 &  & \doi{10.1101/027219} &  &  &  \\
eclair & 2016-01-12 & 2016-05-20 & \doi{10.1093/nar/gkw452} &  &  &  \\
dpt & 2016-02-08 & 2016-08-29 & \doi{10.1038/nmeth.3971} & dpt, wishbone, monocle1 & pseudotime\_correlation, robustness\_unknown & moignard, klein, paul, synthetic\_dpt \\
pseudogp & 2016-04-05 & 2016-11-21 & \doi{10.1371/journal.pcbi.1005212} &  &  &  \\
slicer & 2016-04-09 & 2016-05-23 & \doi{10.1186/s13059-016-0975-3} &  &  &  \\
scell &  & 2016-04-19 & \doi{10.1093/bioinformatics/btw201} &  &  &  \\
wishbone &  & 2016-05-02 & \doi{10.1016/j.cell.2014.04.005} &  &  &  \\
tscan &  & 2016-05-13 & \doi{10.1093/nar/gkw430} & monocle1, tscan, waterfall, scuba, wanderlust & pseudotime\_pos & trapnell, amit, shin \\
scoup &  & 2016-06-08 & \doi{10.1186/s12859-016-1109-3} & scoup, monocle1, tscan & pseudotime\_pis & kouno, moignard, shalek \\
delorean &  & 2016-06-17 & \doi{10.1093/bioinformatics/btw372} &  &  &  \\
raceid\_stemid &  & 2016-06-21 & \doi{10.1016/j.stem.2016.05.010} &  &  &  \\
ouija & 2016-06-23 &  & \doi{10.1101/060442} &  &  &  \\
mpath &  & 2016-06-30 & \doi{10.1038/ncomms11988} &  &  &  \\
celltree &  & 2016-08-13 & \doi{10.1186/s12859-016-1175-6} & monocle1, tscan, celltree & pseudotime\_unknown & trapnell \\
wavecrest &  & 2016-08-17 & \doi{10.1186/s13059-016-1033-x} &  &  &  \\
stemnet &  & 2016-08-25 & \doi{10.1038/ncb3493} &  &  &  \\
scimitar & 2016-10-04 & 2017-01-04 & \doi{10.1142/9789813207813\_0053} & scimitar, monocle1, wanderlust & pseudotime\_correlation & synthetic\_scimitar, synthetic\_scimitar \\
scorpius & 2016-10-07 &  & \doi{10.1101/079509} & scorpius, wanderlust, monocle1, waterfall & pseudotime\_cos, robustness\_cva & schlitzer, buettner, shalek, shalek, shalek, trapnell, kowalczyk, kowalczyk, kowalczyk, kowalczyk \\
scent &  & 2016-10-30 & \doi{10.1038/ncomms15599} & scent, slice, stemid & pseudotime\_wilcox, pseudotime\_auc & chu, trapnell, treutlein \\
slice &  & 2016-12-19 & \doi{10.1093/nar/gkw1278} &  &  &  \\
topslam & 2017-02-13 &  & \doi{10.1101/057778} & monocle1, wishbone, topslam & pseudotime\_correlation & synthetic\_topslam \\
monocle2 & 2017-02-21 & 2017-07-20 & \doi{10.1038/nmeth.4402} & monocle1, monocle2, dpt, wishbone & pseudotime\_correlation, branch\_ari & paul \\
gpfates &  & 2017-03-03 & \doi{10.1126/sciimmunol.aal2192} &  &  &  \\
mfa &  & 2017-03-15 & \doi{10.12688/wellcomeopenres.11087.1} &  &  &  \\
tasic &  & 2017-04-04 & \doi{10.1093/bioinformatics/btx173} &  &  &  \\
somsc & 2017-04-05 &  & \doi{10.1101/124693} &  &  &  \\
slingshot & 2017-04-19 & 2018-06-19 & \doi{10.1186/s12864-018-4772-0} & slingshot, monocle1, monocle2, dpt, tscan & pseudotime\_correlation & synthetic\_splatter, synthetic\_splatter, synthetic\_splatter, synthetic\_splatter, synthetic\_splatter \\
sctda &  & 2017-05-01 & \doi{10.1038/nbt.3854} & sctda, wishbone, slicer, dpt & pseudotime\_correlation & synthetic\_sctda \\
uncurl & 2017-05-31 &  & \doi{10.1101/142398} &  &  &  \\
recat &  & 2017-06-19 & \doi{10.1038/s41467-017-00039-z} & recat, scuba, monocle1, tscan, wishbone, dpt & pseudotime\_correlation, pseudotime\_custom & buettner \\
forks & 2017-06-20 &  & \doi{10.1101/132811} & forks, monocle2, scuba, tscan, waterfall, dpt, gpfates, slicer & pseudotime\_correlation, robustness\_stdev & windram, deng, guo, klein, amit, petropoulos \\
matcher &  & 2017-06-24 & \doi{10.1186/s13059-017-1269-0} & matcher & pseudotime\_correlation & angelmueller, synthetic\_matcher \\
phenopath & 2017-07-06 & 2018-06-23 & \doi{10.1101/159913} &  &  &  \\
hopland &  & 2017-07-12 & \doi{10.1093/bioinformatics/btx232} & hopland, wanderlust, monocle1, topslam, scuba, wishbone, dpt & pseudotime\_correlation & guo, deng, yan, amit, islam, synthetic\_topslam \\
soptsc & 2017-07-26 & 2019-06-20 & \doi{10.1093/nar/gkz204} & soptsc, monocle2, dpt & pseudotime\_correlation & guo, klein, shalek \\
pba &  & 2017-07-30 & \doi{10.1073/pnas.1714723115} &  &  &  \\
brgps & 2017-08-15 &  & \doi{10.1101/167684} & brgps, grandprix, monocle2, scuba, slicer, tscan, wishbone & pseudotime\_correlation & guo, guo \\
wot & 2017-09-27 & 2019-02-07 & \doi{10.1016/j.cell.2019.01.006} &  &  &  \\
treetop & 2017-10-10 &  & \doi{10.1101/200923} &  &  &  \\
paga & 2017-10-27 & 2019-03-19 & \doi{10.1186/s13059-019-1663-x} &  &  &  \\
fateid & 2017-11-11 & 2018-04-09 & \doi{10.1038/nmeth.4662} &  &  &  \\
pseudodynamics & 2017-11-14 &  & \doi{10.1101/219188} &  &  &  \\
pcreode &  & 2017-11-15 & \doi{10.1016/j.cels.2017.10.012} &  &  &  \\
icpsc &  & 2017-11-30 & \doi{10.1038/s41467-017-01860-2} & icpsc, wishbone, monocle2, dpt & pseudotime\_correlation & sun, trapnell, yao \\
grandprix & 2017-12-03 & 2018-07-02 & \doi{10.1101/227843} & delorean, grandprix & pseudotime\_correlation & windram \\
cshmm & 2017-12-03 & 2019-04-30 & \doi{10.1093/bioinformatics/btz296} &  &  &  \\
calista & 2018-01-31 &  & \doi{10.1101/257550} & monocle2, calista, dpt & pseudotime\_correlation & moignard, bargaje, treutlein, chu, synthetic\_calista \\
scepath &  & 2018-02-05 & \doi{10.1093/bioinformatics/bty058} & scepath, monocle1, monocle2, tscan, dpt & pseudotime\_correlation, robustness\_correlation & yan, treutlein, treutlein, trapnell \\
merlot & 2018-02-08 &  & \doi{10.1101/261768} & merlot, dpt, slicer, monocle2, slingshot, tscan & branch\_mi, pseudotime\_correlation & paul, guo, velten, synthetic\_prosstt, synthetic\_prosstt, synthetic\_splatter \\
gpseudorank & 2018-02-08 & 2018-07-25 & \doi{10.1093/bioinformatics/bty664} &  &  &  \\
cellrouter &  & 2018-03-01 & \doi{10.1038/s41467-018-03214-y} & monocle2, dpt, wishbone, waterfall & internal\_autocorrelation & paul, olsson \\
densitypath & 2018-03-05 & 2018-12-07 & \doi{10.1093/bioinformatics/bty1009} & monocle2, wishbone, dpt, densitypath & branch\_ari, pseudotime\_correlation & petropoulos, synthetic\_phate, synthetic\_topslam \\
topographer & 2018-03-23 &  & \doi{10.1101/251207} &  &  &  \\
stream & 2018-04-18 & 2019-04-23 & \doi{10.1038/s41467-019-09670-4} & stream, sctda, wishbone, slicer, monocle2, dpt, tscan, scuba, mpath, gpfates & pseudotime\_correlation & synthetic\_sctda \\
elpigraph & 2018-04-20 &  & arXiv:\href{https://arxiv.org/abs/1804.0758}{1804.0758} &  &  &  \\
urd &  & 2018-04-26 & \doi{10.1126/science.aar3131} &  &  &  \\
celltrails &  & 2018-06-05 & \doi{10.1016/j.celrep.2018.05.002} &  &  &  \\
ddd & 2018-07-12 &  & \doi{10.1101/367789} &  &  &  \\
palantir & 2018-08-05 & 2019-03-21 & \doi{10.1038/s41587-019-0068-4} &  &  &  \\
confess & 2018-09-04 &  & \doi{10.1101/407932} &  &  &  \\
graphddp &  & 2018-09-11 & \doi{10.1038/s41467-018-05988-7} &  &  &  \\
monocle3 &  & 2019-02-28 & \doi{10.1038/s41586-019-0969-x} &  &  &  \\
gpseudoclust & 2019-03-05 &  & \doi{10.1101/567115} & gpseudoclust, monocle2, delorian, slicer & cluster\_ari, cluster\_fmi, cluster\_nmi & sasagawa, shalek, synthetic\_gpseudoclust, synthetic\_gpseudoclust \\
psupertime & 2019-04-29 &  & \doi{10.1101/622001} & monocle2, slingshot, psupertime & pseudotime\_correlation & enge, qiu, petropoulos, li, treutlein \\
cyclum & 2019-05-02 &  & \doi{10.1101/625566} & cyclum, recat & cluster\_accuracy & buettner, mcdavid, mcdavid, mcdavid \\
sinova &  & 2019-05-17 & \doi{10.1016/j.celrep.2016.04.043} &  &  &  \\
\hline
\end{tabularx}
\end{table}
